Sediment transport capacity can vary more in subglacial systems compared to fluvial ones

### 2023-02-17 theory.tex
Updates from Overleaf
Small bits

\ian{check for tense}

\paragraph{Key findings:}
\begin{enumerate}
\item Sediment transport capacity in pressurized subglacial channels  responds differently to changing water discharge in open fluvial channels.
\item Sediment transport capacity varies far more in subglacial channels compared to fluvial ones, even though changing channel width  in fluvial channels accommodates some of the difference.
\item Differences in the sediment transport capacity response to changing water discharge could impact processes at proglacial margins and the interpretation of sediment transport records from glaciers. 
\end{enumerate}

@@

  However, in response to changing water discharge, the behavior of subglacial channels diverges from fluvial ones.
  % In fluvial channels, the channel width and height may evolve with changing flow conditions, but the underlying channel geometry responds on a far longer timescale than water discharge of the water discharge \citep[Figure~\ref{fig:cartoon}\,b and d; e.g. ][]{church2006}.
  In subglacial channels, however, the water is pressurized in a subglacial channel and the size of this channel responds to the roughness of the channel walls and the velocity of the water flowing through the channel \citep[Figure~\ref{fig:cartoon}\,a and c; e.g. ][]{rothlisberger1972}.
  Here, diurnal variations in water flux, for instance, may cause water to flow through a conduit whose size does not respond to the same timescales as the increase in water flux \citep{rothlisberger1972}.
  With respect to sediment transport, this means that increased water discharge in subglacial channels will result in greater water velocity and thus sediment transport capacity.
  At the same time, the fixed channel width in subglacial channels means that the channel width over which to mobilize sediment does not evolve over the same timescales as water discharge \citep[e.g.][]{werder2010}.

@@

  Next, the setup is run $25,000$ times with randomly selected channel geometries ($k=(2,6)$, $\beta=(\frac{\pi}{10},\pi)$), friction factors ($f_i=(0.001,20)$, $f_p=(0.001,20)$) and water discharges (maximum and minimum values selected between $5$ to $500$ \,\mmma). $Q_w^*$ in Equation ~\ref{eq:DW} is determined as the midpoint in the water discharge range. Parameter combinations are only selected if their maximum and minimum velocities are between $0.2$ and $3$\,\unit{m}\,\unit{s}$^{-1}$, resulting in roughly $10,600$ adequate parameter combinations.

  \begin{center}
    \begin{figure}[H]
      \includegraphics[width=0.9\linewidth]{multi_run.pdf}
      \caption{Ratio of range of a variable $\Gamma$ for glacial and fluvial systems for water velocity  (a), shear-stress (b), and  width-integrated shear-stress (c) in response to mean water discharge,
        Range  $\chi$ for each variable is given in Equation~\ref{eq:var}. Values greater (less) than $1$ shown in white (gray) are parameter combinations that vary more in a subglacial (fluvial) system. Plots (d-f) show the response of variable $\Gamma$ to the variability in water discharge ($\chi(Q_w)$) in the parameter combination. 
        Histograms (g-i) show the frequency of variations across variables in the parameterization.}
      \label{fig:range}
    \end{figure}
  \end{center}

@@

  The subglacial parameterization assumes that the friction factor does not evolve between the formation of the channel and the sediment transport conditions examined.
  The role of this effect is uncertain, and most numerical models assume a steady friction factor \citep[e.g.][]{werder2013}. 
  However, the friction factor has been documented to evolve in time \citep{pohle2022}.
  Furthermore, both subglacial and fluvial parameterizations here assume that the distribution of velocity and shear stress of water across the channel bed is homogeneous (Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}).

  The subglacial parameterization also assumes that subglacial water is pressurized, which is a commonly held assumption in subglacial hydrology \citep[e.g.][]{gimbert2016}.
  While the subglacial channel geometry need not be in a steady state for the water to remain pressurized, the subglacial parameterization assumes that the channel geometry must vary minimally over the timescales of varying  water discharge \citep[e.g.][]{nanni2020}.
  Note that in R-channel formulations with changing melt input, the variability in sediment transport capacity could be somewhat smaller than the parameterizations here.
  Here, peaks in the hydrograph would occur as the R-channel is opening due to pressure melt, thus causing a decrease in velocity compared to the results here \citep{rothlisberger1972}.
  A similar effect could occur with creep closure of the subglacial channel at the hydrograph minima.

  The fluvial channel parameterization (Section~\ref{sect:fluv}) assumes that sediment transported across all channel widths, as opposed bankfull width when most sediment transport occurs \citep{wolman1960}.
  As a result, the variability in the fluvial system across the range of discharge values could be underestimated in the parameterization (Figure~\ref{fig:range}).

@@

  The response of sediment discharge caused by the greater variability in sediment transport in subglacial channels will be greatest in completely transport-limited regimes \citep{kasmalkar2019}.
  In supply-limited systems, the changing sediment transport capacity (Figure~\ref{fig:model_outs}) would have a minimal effect on sediment discharge due to the absence of sediment available for transport \citep{delaney2019}.
  Yet, reaching the threshold of motion for sediment more frequently may result in additional sediment transport and the tendency for glaciers to transport sediment in a supply-limited case \citep[e.g.][]{herman2015}.

  The effect of pressurized channel flow could be most pronounced during precipitation or flood events when subglacial conduits have not adapted to the water discharge flowing through them, resulting in a very large increase in sediment transport capacity \citep{delaney2019}.
  Precipitation events have been documented to result in exceptionally high sediment discharge from glaciers \citep{cowan1988,delaney2018}.
  Increased sediment discharge may result from changes to sediment access below the glacier, yet rapid increases in water velocity from water flowing through a small conduit, not conditioned by the amount of water flow, will cause a large increase in sediment transport capacity (Section~\ref{sect:sub_mode}).

  Increased variability in sediment transport also impacts sediment export from glacier margins into proglacial areas \citep[e.g.][]{lane2016,delaney2017,perolo2018}.
  Here, in the transition to open-channel flow as the water leaves the glacier, velocities at the highest water discharges could generally decrease (Figure~\ref{fig:model_outs}, blue and green horizontal bars), potentially resulting in sediment deposition once sediment enters the proglacial area.
 Note that the transition from pressurized to open channel flow could occur above the glacier's terminus, when cavity closure rates are reduced and water can flow freely \citep{egli2021b}, resulting in intermittent deposition and remobilization of sediment below the glacier terminus \citep{perolo2018}.

@@

   Records of sediment discharge have been used to establish the relationship, or lack thereof, between sediment discharge and climate in glacial systems \citep[e.g.][]{koppes2009a,willenbring2016,mariotti2021}.
  The variability in subglacial sediment transport capacity presented here applies most generally to short-time scales responsible for the size of subglacial channels.
  Yet, variability in sediment transport from glaciers may add to the already large amount of noise in the fluvial transport of sediment from the glacier to the sink  \citep{castletort2003,jerolmack2010,romans2016}, should the glacier be land-terminating with a proglacial river connecting the glacier to the sink of sediment.
   Therefore, identifying climatic signals in sediment transport from transport-limited glaciers may require higher thresholds of climatic perturbation compared to fluvial systems \citep{tofelde2021}, given the amount of noise present.
  For instance, sediment discharge records from two glaciers in the Swiss Alps show that $40$ to $50$\% of a season's sediment discharge occurs when water discharge is below the $75^{\mathrm{th}}$ quantile of the season \citep{delaney2018}, possibly because sediment transport capacity relies on the size of the glacier conduit, in addition to water discharge (Section~\ref{sect:sub_mode}).
  This suggests that a larger perturbation in water discharge might be needed to significantly alter the sediment discharge from the glacier compared to a fluvial system, should the subglacial sediment transport be in a largely transport-limited regime.

  Additional processes and erosional hiatuses may further complicate signals of sediment discharge from glaciers in response to climate, compared to fluvial systems \citep{jansson2005,ganti2016}. 
  If a glacier expels sediment in a transport-limited regime, the pressurized nature of subglacial water flow means sediment discharge capacity responds to the thickness of the ice, controlling the closure of the channel, and the surface slope of the glacier \citep{rothlisberger1972,shreve1972}, controlling water velocity, in addition to water discharge (Section~\ref{sect:sub_mode}).
  Conversely, sediment transport in most fluvial systems increases with water discharge and bed slope \citep[Section~\ref{sect:fluv}; e.g.][]{muller1968,whipple1999,wong2006}. 
  Furthermore, if the glacier is in the supply-limited case, sediment discharge record will also represents additional processes of sliding  \citep{herman2015,seguinot2021}, adding additional processes represented in the sediment discharge signal \citep{delaney2019}.

@@

  Divergent processes occur in sediment transport capacity's response to water discharge that exists between subglacial and fluvial systems due to their fixed and variable channel geometry, respectively.
  Parameterization of these processes shows that variability in water velocity and shear stress in response to variable water discharge is much higher in subglacial channels than in fluvial ones.
  The variability in shear stress is reduced when the evolving channel width is accounted for in fluvial systems, yet, the variability in width-integrated shear stress is still higher in subglacial systems.
  When shear stress and width integrated shear stress are linked to sediment transport capacity \citep[e.g.][]{shields1936,meyer1948,engelund1967}, sediment transport capacity can vary far more in subglacial systems compared to fluvial ones.

  A subglacial channel's greater variability in sediment transport capacity is a process that limits geomorphic comparison with fluvial systems, given impacts on the mobilization, transport, and deposition of sediment below glaciers.
  This variability in sediment transport capacity could cause sediment deposition for sediment transport in glacier margins, where water flow transitions from pressurized subglacial flow to open-channel and fluvial.
  Sediment transport in glacier systems responds to a large number of factors such as ice thickness and glacier surface slope, which may respond to climate differently than water discharge. 
  The inconsistent response of subglacial and fluvial systems to water discharge could adds additional noise to sediment transport signals in glacial systems compared to fluvial ones.

@@

\bibliographystyle{apa}
\bibliography{Paperlib}

@@

>>>>

        \paragraph{Key findings:}
        \begin{enumerate}
        \item Sediment transport capacity varies more in pressurized subglacial channels compared to open fluvial ones, while changing channel width  in fluvial channels accommodates some of the difference.
        \item Differences in the sediment transport capacity response to changing water discharge could impact processes at proglacial margins and the interpretation of sediment transport records from glaciers. 
        \end{enumerate}

@@

          However, in response to changing water discharge, the behavior of subglacial channels diverges from fluvial ones.
          In subglacial channels, however, the water is pressurized in a subglacial channel and the size of this channel responds to the roughness of the channel walls and the velocity of the water flowing through the channel \citep[Figure~\ref{fig:cartoon}\,a and c; e.g. ][]{rothlisberger1972}.
          Here, diurnal variations in water flux, for instance, may cause water to flow through a conduit whose size does not respond to the same timescales as the increase in water flux \citep{rothlisberger1972}.
          With respect to sediment transport, this means that increased water discharge in subglacial channels will result in greater water velocity and thus sediment transport capacity.
          At the same time, the fixed channel width in subglacial channels means that the channel width over which to mobilize sediment does not evolve over the same timescales as water discharge \citep[e.g.][]{werder2010}.

@@

          Next, the parameterizations are run $25,000$ times with randomly selected channel geometries ($k=(2,6)$, $\beta=(\frac{\pi}{10},\pi)$), friction factors ($f_i=(0.001,20)$, $f_p=(0.001,20)$) and water discharges (maximum and minimum values selected between $5$ to $500$ \,\mmma). $Q_w^*$ in Equation ~\ref{eq:DW} is determined as the midpoint in the water discharge range. Parameter combinations are only selected if their maximum and minimum velocities are between $0.2$ and $3$\,\unit{m}\,\unit{s}$^{-1}$, resulting in roughly $10,600$ adequate parameter combinations.

          \begin{center}
            \begin{figure}[H]
              \includegraphics[width=0.9\linewidth]{multi_run.pdf}
              \caption{Ratio of range of a variable $\Gamma$ for glacial and fluvial systems for water velocity  (a), shear-stress (b), and  width-integrated shear-stress (c) in response to mean water discharge,
                Range  $\chi$ for each variable is given in Equation~\ref{eq:var}. Values greater (less) than $1$ shown in white (gray) are parameter combinations that vary more in a subglacial (fluvial) system. Plots (d-f) show the response of variable $\Gamma$ to the variability in water discharge ($\chi(Q_w)$) in the parameter combination. 
                Histograms (g-i) show the frequency of relative variations across variables in the parameterization.}
              \label{fig:range}
            \end{figure}
          \end{center}

@@

          The subglacial parameterization also assumes that subglacial water is pressurized, which is a commonly held assumption in subglacial hydrology \citep[e.g.][]{gimbert2016}.
          While the subglacial channel geometry need not be in a steady state for the water to remain pressurized, the subglacial parameterization assumes that the channel geometry must vary minimally over the timescales of varying  water discharge \citep[e.g.][]{nanni2020}.

          In R-channel formulations with changing melt input, the variability in sediment transport capacity could be somewhat smaller than the parameterizations here.
          Here, peaks in the hydrograph could cause additional opening of the R-channel  due to pressure melt, thus causing a decrease in velocity compared to the results here \citep{rothlisberger1972}.
          A similar effect could occur with creep closure of the subglacial channel at the hydrograph minima.
          Additionally, at high water pressures, glacier uplift might change the access of subglacial water to sediment \citep{iken1993}.

          Both subglacial and fluvial parameterizations here assume that the distribution of velocity and shear stress of water across the channel bed is homogeneous (Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}). 
          The fluvial parameterization assumes that sediment is transported across all channel widths, as opposed bankfull width when much sediment transport occurs \citep{wolman1960}.
          As a result, the variability in the fluvial system across the range of discharge values could be underestimated in the parameterization (Figure~\ref{fig:range}).

@@

          The response of sediment discharge caused by the greater variability in sediment transport in subglacial channels will be greatest in completely transport-limited regimes \citep{kasmalkar2019}.
          In supply-limited systems, the changing sediment transport capacity (Figure~\ref{fig:model_outs}) would have a minimal effect on sediment discharge due to the absence of sediment available for transport \citep{delaney2019}.
          Yet, reaching the threshold of motion for sediment more frequently may result in additional sediment transport and the tendency for glaciers to transport sediment in a supply-limited regime \citep[e.g.][]{herman2015}.

          The effect of pressurized channel flow could be most pronounced during precipitation or flood events when subglacial conduits have not adapted to the water discharge flowing through them, resulting in a very large increase in sediment transport capacity and discharge \citep[e.g.][]{cowan1988,delaney2019}.
         Increased sediment discharge may result from changes to sediment access below the glacier, yet rapid increases in water velocity from water flowing through a small conduit, not conditioned by the amount of water flow, will cause a large increase in sediment transport capacity (Section~\ref{sect:sub_mode}).

          Increased variability in sediment transport also impacts sediment export from glacier margins into proglacial areas \citep[e.g.][]{delaney2017,perolo2018}.
          Here, in the transition to open-channel flow as the water leaves the glacier, velocities at the highest water discharges could generally decrease (Figure~\ref{fig:model_outs}, blue and green horizontal bars), potentially resulting in sediment deposition once sediment enters the proglacial area.
         Note that the transition from pressurized to open channel flow could occur above the glacier's terminus, when cavity closure rates are reduced and water can flow freely \citep{egli2021b}, resulting in intermittent deposition and remobilization of sediment below the glacier terminus \citep{perolo2018}.

@@

                    Additional processes and erosional hiatuses may further complicate signals of sediment discharge from glaciers in response to climate, compared to fluvial systems \citep{jansson2005,ganti2016}. 
    For instance, if a glacier expels sediment in a transport-limited regime, the pressurized nature of subglacial water flow means sediment discharge capacity responds to the thickness of the ice, controlling the closure of the channel, and the surface slope of the glacier \citep{rothlisberger1972,shreve1972}, controlling water velocity, in addition to water discharge (Section~\ref{sect:sub_mode}).
          Conversely, sediment transport in most fluvial systems increases with water discharge and bed slope \citep[Section~\ref{sect:fluv}; e.g.][]{muller1968,whipple1999,wong2006}. 
          Furthermore, if the glacier is in the supply-limited case, the sediment discharge record will also represent additional processes of sliding  \citep{herman2015,seguinot2021}, adding additional processes represented in the sediment discharge signal \citep{delaney2019}.

           Records of sediment discharge have been used to establish the relationship, or lack thereof, between sediment discharge and climate in glacial systems \citep[e.g.][]{koppes2009a,willenbring2016,mariotti2021}.
          The variability in subglacial sediment transport capacity presented here applies most generally to short-time scales responsible for the size of subglacial channels.
          Yet, variability in sediment transport from glaciers may add to the already large amount of noise in the fluvial transport of sediment from the glacier to the sink  \citep{castletort2003,jerolmack2010,romans2016}, especially if the glacier is land-terminating with a proglacial river connecting the glacier to the sink of sediment.
           Therefore, identifying climatic signals in sediment transport from transport-limited glaciers may require higher thresholds of climatic perturbation compared to fluvial systems \citep{tofelde2021}, given the amount of noise present.
          For instance, sediment discharge records from two glaciers in the Swiss Alps show that $40$ to $50$\% of a season's sediment discharge occurs when water discharge is below the $75^{\mathrm{th}}$ quantile of the season \citep{delaney2018}, possibly because sediment transport capacity relies on the size of the glacier conduit, in addition to water discharge (Section~\ref{sect:sub_mode}).
          This suggests that a larger climate perturbation might be needed to significantly alter the sediment discharge capacity from the glaciers compared to fluvial systems, to overcome the increased variability in sediment transport capacity.

@@

            Application of parameterizations for subglacial and fluvial water flow shows that divergent processes occur in sediment transport capacity's response to water discharge, due to their fixed and variable channel geometry, respectively.
            Results show that the impact of variable water discharge on in water velocity and shear stress, first-order controls on sediment mobilization, is much higher in subglacial channels than in fluvial ones.
              The variability in shear stress is reduced when the evolving channel width is accounted for in fluvial systems.
    Even so, variability in shear stress across a channel's width is still higher in subglacial channels, compared to fluvial ones.

    These divergent characteristics between glacial and fluvial channels show that records of sediment transport downstream of glaciers capture multiple processes together.
    Such processes may compound at glacier margins, where water flow transitions from pressurized subglacial flow to open-channel flow.
      The inconsistent response of subglacial and fluvial systems to water discharge could add  noise to sediment transport signals in glacial systems compared to fluvial ones.
        In addition, variations in sediment transport capacity in glacier systems respond to a large number of factors such as channel size, ice thickness, and glacier surface slope, which may react to climate forcing differently than water discharge alone. 
      The differences in the systems' response to changing water discharge should be considered when examining sediment transport processes in glacierized catchments.

@@

        \bibliographystyle{apalike}
        \bibliography{Paperlib.bib}

@@

Qbs_out Qws_out
